---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

* Set the environment

```{r}
#Test Scripts
##Solution Environment
library(dplyr)
library(data.table)
library(lubridate)
library(ggplot2)

```

1. Extract dataset as Data Table

```{r}
mad <- fread("activity/activity.csv")
```

## What is mean total number of steps taken per day?

2. Calculate the total number of steps taken per day
```{r}
by_date <- group_by(mad, date)
madTotalSteps <- summarise(by_date, totStepsPerDay = sum(steps, na.rm = FALSE))
```

3. Make a histogram of the total number of steps taken each day

```{r}
hist(madTotalSteps$totStepsPerDay, main="Daily Steps Frequency", xlab="Total Steps", breaks=nrow(madTotalSteps))
```

4. Calculate and report the mean and median of the total number of steps taken per day

```{r}
madMeanSteps <- summarise(by_date, meanSteps = mean(steps, na.rm = TRUE))
madMedSteps <- summarise(by_date, medSteps = median(steps, na.rm = TRUE))
```

## What is the average daily activity pattern?

* Make a time series plot (i.e.l) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
by_int <- group_by(mad, interval)
madMeanIntSteps <- summarise(by_int, intMeanSteps = mean(steps, na.rm=TRUE))

ggplot(madMeanIntSteps, aes(x = interval , y = intMeanSteps)) + geom_line(color="blue", size=1) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")
```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxInt <- arrange(madMeanIntSteps, desc(intMeanSteps)) %>% slice(1)
```

## Imputing missing values

* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nrow(mad[is.na(steps),])
```

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

1. Use the mean to fill in the NA values in the dataset

```{r}
mad[is.na(steps), "steps"] <- mad[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps")]
```

2. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
data.table::fwrite(x = mad, file = "tidyActivity.csv", quote = FALSE)
```

3. Extract the clean dataset

```{r}
madClean <- fread("tidyActivity.csv")

```

4. Make a histogram of the total number of steps taken each day

```{r}
by_date <- group_by(madClean, date)
mcTotalSteps <- summarise(by_date, totStepsPerDay = sum(steps, na.rm = FALSE))
hist(mcTotalSteps$totStepsPerDay, main="Daily Steps Frequency", xlab="Total Steps", breaks=nrow(mcTotalSteps))

```

5. Calculate and report the mean and median total number of steps taken per day. 

```{r}
mcMeanTotSteps <- mean(mcTotalSteps$totStepsPerDay, na.rm = FALSE)
mcMedTotSteps <- median(mcTotalSteps$totStepsPerDay, na.rm = FALSE)
```


6. Do these values differ from the estimates from the first part of the assignment?

```{r}
madMeanTotSteps <- mean(madTotalSteps$totStepsPerDay, na.rm = TRUE)
madMedianTotSteps <- median(madTotalSteps$totStepsPerDay, na.rm = TRUE)

mmLabel <- c("Mean", "Median")
oldMM <- c(madMeanTotSteps, madMedianTotSteps)
newMM <- c(mcMeanTotSteps, mcMedTotSteps)

mmComparison <- data.table(mmLabel, oldMM, newMM)
```

* It can be seen that the values do differ after inputting the missing variables as Mean

```{r}
mmComparison
```

* What is the impact of imputing missing data on the estimates of the total daily number of steps?

## It increases the frequency of steps per day

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
mad[, date := as.POSIXct(date, format = "%Y-%m-%d")]
mad[, `Day of Week`:= weekdays(x = date)]
mad[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
mad[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
mad[, `weekday or weekend` := as.factor(`weekday or weekend`)]
```


2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
mad[is.na(steps), "steps"] <- mad[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
madInterval <- mad[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)] 
ggplot(madInterval , aes(x = interval , y = steps, color=`weekday or weekend`)) + geom_line() + labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") + facet_wrap(~`weekday or weekend` , ncol = 1, nrow=2)
```

